Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2023-05-30 paper/ms.tex
Further changes to paper

In particular, we find that the presence of a significant gas phase in a
barred galaxy can arrest the process by which the dark matter halo brakes the
bar. The constant positive torque on the bar by the gas causes the bar's
resonance to reside in regions which have become depopulated. In the halo wake
picture, this is equivalent to no new material being available to reinforce the
wake. We show that this occurs in Milky Way-like galaxies with gas fractions as
low as about $5\%$.

@@

\begin{figure}
    \centering
    \includegraphics[width=9cm]{fig/Rb_A2.pdf}
    \caption{The evolution in bar length, mass and strength. The \textit{upper panel}
    shows the evolution of the bar length. In the \Nbody{} case, the bar
    lengthens. This occurs because as the pattern speed drops, bar-like orbits
    at larger radii are possible. Stars are captured on these orbits,
    lengthening the bar. This process does not occur in the \SMUGGLE{} cases
    since the bar pattern speed is not decreasing, and therefore the bar length
    remains constant. The bar mass, shown in the \textit{middle panel}, is
    increasing in the \Nbody{} case as the bar grows. It is decreasing in the
    \SMUGGLE{} case, indicating {\bf mass redistribution without a change in patttern speed}. The bar strength, shown in
    the \textit{lower panel} is measured as the maximum of the second Fourier
    component divided by the zeroth Fourier component. We see that in the
    \Nbody{} case (blue) the bar strength increases with time, consistent with
    previous results showing that the bar strength increases as bars slow down.
    In the \SMUGGLE{} case (orange), we see that the bar strength slightly
    decreasing with time. This is also consistent with the expected relation
    between pattern speed and strength since the bar in this case is not slowing
    down.}
    \label{fig:strength}
\end{figure}
\subsection{Initial Conditions}
The initial setup of the galactic disk used in this work follows closely the
GALAKOS model \citep{2020ApJ...890..117D}, which uses a modified version of the
\texttt{MakeNewDisk} code \citep{2005MNRAS.361..776S}. The GALAKOS model has
three components - a radially exponential and vertically isothermal stellar
disk, and a stellar bulge and dark matter halo following a Hernquist profile
\citep{1990ApJ...356..359H}. All \Nbody{} runs in this work used the same setup
parameters as the GALAKOS disk, more details of which can be found in the
original paper.

The addition of the gas phase was done as follows. The version of
\texttt{MakeNewDisk} used for the original GALAKOS model can generate a gas disk
which is radially exponential and in vertical gravito-hydrodynamic balance. We
modified the radial profile of this code in order to allow us to generate a disk
with a constant surface density within some cut-off radius, and then
exponentially declining beyond that radius with the scale-length of the stellar
disk. Our fiducial model used an initial surface density of
$20\,M_{\odot}/\textrm{pc}^2$ and a cut-off radius of $9.3\,\textrm{kpc}$. The
initial gas disk is generated with a temperature of $10^4\,\textrm{K}$ and solar
metallicity.

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/halo_trapped.pdf}
    \caption{The mass fraction within two disk scale lengths
    ($\sim5.3\,\textrm{kpc}$) trapped in the \Nbody{} and \SMUGGLE{} runs. As
    the dark matter halo torques the bar, material from the halo is trapped on
    bar-like orbits. In the \Nbody{} case, the trapped fraction increases with
    time, indicating the torquing process is active and the bar is unlocked. In
    the \SMUGGLE{} case, the trapped fraction is nearly constant with time,
    indicating the torquing process is inactive and the bar is locked.}
    \label{fig:htrap}
\end{figure}

@@

\subsection{Varying Initial Gas Fractions}
We performed a test in which we varied the initial gas fraction of the
disk. In our fiducial run, we set the surface density of the gas disk from
$4\,\textrm{kpc}$ to $\sim9.3\,\textrm{kpc}$ to be $20\,\Msun/\textrm{pc}^2$.
We also ran with surface densities of $15$, $10$, and
$5\,\Msun/\textrm{pc}^2$. These correspond to initial gas fractions of
approximately $16\%$, $10\%$, $7\%$, and $4\%$. The pattern speed evolution is
shown in Fig.~\ref{fig:fgas} We find that the bar in disks with
initial surface densities of $20$, $15$, and $10\,\Msun/\textrm{pc}^2$ evolve
with a constant pattern speed while the bar in a disk with initial surface
density of $5\,\Msun/\textrm{pc}^2$ slows down at a similar rate to the \Nbody{} case.

After $5\,\textrm{Gyr}$, the $10\,\Msun/\textrm{pc}^2$ simulation has a gas
fraction of $5.7\%$, but still exhibits constant pattern speed behavior. As a
result, we conclude that for the disk, bar, and halo properties considered in
this work, a gas fraction of only approximately $5\%$ is necessary in order for
the proposed stabilizing mechanism to operate.

\subsection{Smooth Interstellar Medium}
We performed a simulation of the same disk but with a simpler model of the
interstellar medium \citep{2003MNRAS.339..289S}, closer to standard methods used
in cosmological simulations of galaxy formation and described in more detail in
Section~\ref{sec:methods}. The result of this test is presented in
Fig.~\ref{fig:GFM}. We find that the pattern speed evolution is nearly the same
in this case, and so conclude that our result is not sensitive to the details of
the model for the interstellar medium.

@@

\citet{2013MNRAS.429.1949A, 2014MNRAS.438L..81A} find stable evolution in
\Rot{} for a triaxial halo over $10\,\textrm{Gyr}$ with initial (final) gas
fractions of $100\%$ ($7\%$), $75\%$ ($6\%$), and $50\%$ ($5\%$). For a model
with $20\%$ ($3\%$) gas fraction they find an increasing \Rot{}. In their model
with a spherical halo they always find increasing \Rot{}, contrary to the
present work. We note there is a structural difference in their dark matter
halo. They use a cored isothermal sphere as opposed to our Hernquist halo. It is
not clear to us the impact this would have on the expected torque from the halo
(i.e., the velocity structure of their halo may allow for more efficient capture
and thus stronger torquing). These authors make no mention of the process by
which the halo braking process is arrested, as we propose in the present paper.

@@

\begin{acknowledgments}
  We would like to thank Greg~L. Bryan, Neal~J. Evans, Drummond~B. Fielding,
  Keith Hawkins, Jason~A.~S. Hunt, Sarah~M.~R. Jeffreson, Kathryn~V. Johnston,
  Peter~M.~W. Kalberla, Jürgen Kerp, Julio~F. Navarro, Dylan Nelson, Suchira
  Sarkar, Joshua~S. Speagle, Martin~D. Weinberg, and Yanfei Zou for helpful
  discussions. A.B. would like to thank Todd Phillips for helpful discussions.
  Some computations in this paper were run on the FASRC Cannon cluster supported
  by the FAS Division of Science Research Computing Group at Harvard University.
  Resources supporting this work were also provided by the NASA High-End
  Computing (HEC) Program through the NASA Advanced Supercomputing (NAS)
  Division at Ames Research Center. This research has made use of the Spanish
  Virtual Observatory (https://svo.cab.inta-csic.es) project funded by
  MCIN/AEI/10.13039/501100011033/ through grant PID2020-112949GB-I00. A.B. was
  supported by the Future Investigators in NASA Earth and Space Science and
  Technology (FINESST) award number 80NSSC20K1536 during the completion of this
  work. E.D. was partially supported by HST grants HST-AR-16363.001 and
  HST-AR-16602.006-A and by NASA Award NASA 80NSSC22K0761. J.Q. acknowledges
  support from NSF grant AST-2008490. L.V.S. is grateful for financial support
  from NASA ATP 80NSSC20K0566, NSF AST 1817233 and NSF CAREER 1945310 grants.
  P.T. acknowledges support from NSF grant AST-1909933, AST-2008490, and NASA
  ATP Grant 80NSSC20K0502. M.V. acknowledges support through NASA ATP
  19-ATP19-0019, 19-ATP19-0020, 19-ATP19-0167, and NSF grants AST-1814053,
  AST-1814259, AST-1909831, AST-2007355 and AST-2107724.
\end{acknowledgments}

@@

\section{Stars Instead of Gas}
In the SMUGGLE model considered in this work, we instantaneously added gas to
the $N$-body system after $1.5\,\textrm{Gyr}$ of evolution. One might wonder if
this sudden change to the potential is responsible for the stable pattern speed
evolution. To test whether this is the case, we added mass to the system in the
same way we did for the SMUGGLE model, but using collisionless particles instead
of gas. The result of this experiment is shown in Fig.~\ref{fig:ps-star}. While
there is an offset compared to the pure $N$-body case, we see that the pattern
speed evolution is broadly consistent with a declining pattern speed. This
indicates that the gas phase is responsible for the stable pattern speed.
\section{Semi-Analytic Model Parameters}
\label{app:sam}
Our semi-analytic model consisted of a three-component bar-disk-halo system. We
describe here the parameters we chose for these components. The parameters of
the disk and halo were chosen to match closely what we used in our fiducial
simulations. The system can thus be understood as being roughly similar to the
Milky Way, though no careful analysis has been performed to ensure the closest
match possible.

>>>>

In particular, we find that the presence of a gas phase in a barred
galaxy can arrest the process by which the dark matter halo brakes the bar. The
constant positive torque on the bar by the gas causes the bar's resonance to
reside in regions which have become depopulated. In the halo wake picture, this
is equivalent to no new material being available to reinforce the wake. We show
that this occurs in {\bf a Milky Way-like disk} with gas fractions as low as
about $5\%$.

@@

\begin{figure}
    \centering
    \includegraphics[width=9cm]{fig/Rb_A2.pdf}
    \caption{The evolution in bar length, mass and strength. The \textit{upper panel}
    shows the evolution of the bar length. In the \Nbody{} case, the bar
    lengthens. This occurs because as the pattern speed drops, bar-like orbits
    at larger radii are possible. Stars are captured on these orbits,
    lengthening the bar. This process does not occur in the \SMUGGLE{} cases
    since the bar pattern speed is not decreasing, and therefore the bar length
    remains constant. The bar mass, shown in the \textit{middle panel}, is
    increasing in the \Nbody{} case as the bar grows. It is decreasing in the
    \SMUGGLE{} case, indicating {\bf mass redistribution without a change in
    pattern speed}. The bar strength, shown in the \textit{lower panel} is
    measured as the maximum of the second Fourier component divided by the
    zeroth Fourier component. We see that in the \Nbody{} case (blue) the bar
    strength increases with time, consistent with previous results showing that
    the bar strength increases as bars slow down. In the \SMUGGLE{} case
    (orange), we see that the bar strength slightly decreasing with time. This
    is also consistent with the expected relation between pattern speed and
    strength since the bar in this case is not slowing down.}
    \label{fig:strength}
\end{figure}
\subsection{Initial Conditions}
The initial setup of the galactic disk used in this work follows closely the
GALAKOS model \citep{2020ApJ...890..117D}, which uses a modified version of the
\texttt{MakeNewDisk} code \citep{2005MNRAS.361..776S}. The GALAKOS model has
three components - a radially exponential and vertically isothermal stellar
disk, and a stellar bulge and dark matter halo following a Hernquist profile
\citep{1990ApJ...356..359H}. All \Nbody{} runs in this work used the same setup
parameters as the GALAKOS disk, more details of which can be found in the
original paper.

The addition of the gas phase was done as follows. The version of
\texttt{MakeNewDisk} used for the original GALAKOS model can generate a gas disk
which is radially exponential and in vertical gravito-hydrodynamic balance. We
modified the radial profile of this code in order to allow us to generate a disk
with a constant surface density within some cut-off radius, and then
exponentially declining beyond that radius with the scale-length of the stellar
disk. Our fiducial model used an initial surface density of
$20\,M_{\odot}/\textrm{pc}^2$ and a cut-off radius of $9.3\,\textrm{kpc}$. {\bf
This corresponds to an initial gas fraction of $\sim16\%$.} The initial gas disk
is generated with a temperature of $10^4\,\textrm{K}$ and solar metallicity.

@@

{\bf Our method for initializing gas was arrived at after numerous attempts to
include enough gas in the simulation to be compatible with the Milky Way while
also not destroying the bar. For example, we tried evolving an exponential gas
disk adiabatically with the barred disk (i.e., no cooling, star formation, or
feedback). Because there is no mechanism to remove gas from the central region,
a large pileup of gas forms in the center.\footnote{{\bf We noticed that the bar
slows down in this adiabatic model which lacks a mechanism to remove gas from
the central region. \citet{2007ApJ...666..189B} also noted slowdown behavior in
models which lacked central gas removal. It appears that gas removal from the
center (in our model due to star formation) is necessary to stabilize a bar's
pattern speed.}} Once cooling is turned on, there is a sudden collapse of gas to
the center, destroying the bar. We don't believe our method is the only nor even
the best way to include gas. One advantage of our method, though, is that it
approximates the expectation that in the bar region the gas surface density
should be significantly reduced \citep[e.g.][]{1993RPPh...56..173S}}

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/halo_trapped.pdf}
    \caption{The {\bf halo} mass fraction within two disk scale lengths
    ($\sim5.3\,\textrm{kpc}$) trapped in the \Nbody{} and \SMUGGLE{} runs. As
    the dark matter halo torques the bar, material from the halo is trapped on
    bar-like orbits. In the \Nbody{} case, the trapped fraction increases with
    time, indicating the torquing process is active and the bar is unlocked. In
    the \SMUGGLE{} case, the trapped fraction is nearly constant with time,
    indicating the torquing process is inactive and the bar is locked.}
    \label{fig:htrap}
\end{figure}

@@

\subsection{Varying Initial Gas Fractions}
We performed a test in which we varied the initial gas fraction of the
disk. In our fiducial run, we set the surface density of the gas disk from
$4\,\textrm{kpc}$ to $\sim9.3\,\textrm{kpc}$ to be $20\,\Msun/\textrm{pc}^2$.
We also ran with surface densities of $15$, $10$, and
$5\,\Msun/\textrm{pc}^2$. These correspond to initial gas fractions of
approximately $16\%$, $10\%$, $7\%$, and $4\%$. The pattern speed evolution is
shown in Fig.~\ref{fig:fgas}. We find that the bar in disks with
initial surface densities of $20$, $15$, and $10\,\Msun/\textrm{pc}^2$ evolve
with a constant pattern speed while the bar in a disk with initial surface
density of $5\,\Msun/\textrm{pc}^2$ slows down at a similar rate to the \Nbody{} case.

After $5\,\textrm{Gyr}$, the $10\,\Msun/\textrm{pc}^2$ simulation has a gas
fraction of $5.7\%$, but still exhibits constant pattern speed behavior. As a
result, we conclude that for the disk, bar, and halo properties considered in
this work, a gas fraction of only approximately $5\%$ is necessary in order for
the proposed stabilizing mechanism to operate. {\bf We stress that this gas
fraction threshold is only for the system considered in this work. Systems with
different structural parameters may require a different threshold. For example,
\citet{2015MNRAS.454.3166A} find slowdown behavior in a barred disk $10$ times
less massive than ours with a gas fraction of $10\%$. To make things more
complicated, \citet{2010ApJ...719.1470V} find that the gas fraction cutoff
varies with the softening length used. Determining exactly how the gas fraction
cutoff varies with these considerations deserves further attention.}

@@

\citet{2013MNRAS.429.1949A, 2014MNRAS.438L..81A} find stable evolution in \Rot{}
for a triaxial halo over $10\,\textrm{Gyr}$ with initial (final) gas fractions
of $100\%$ ($7\%$), $75\%$ ($6\%$), and $50\%$ ($5\%$). For a model with $20\%$
($3\%$) gas fraction they find an increasing \Rot{}. In their model with a
spherical halo they always find increasing \Rot{}, contrary to the present work.
We note there is a structural difference in their dark matter halo. They use a
cored isothermal sphere as opposed to our Hernquist halo. It is not clear to us
the impact this would have on the expected torque from the halo (i.e., the
velocity structure of their halo may allow for more efficient capture and thus
stronger torquing). {\bf The bar locking process (or something similar) which we
propose in this paper is not mentioned by these authors.}

@@

{\bf Snapshots at $500\,\textrm{Myr}$ cadence are publicly available at
\url{https://drive.google.com/drive/folders/1nZF7mZ98T0QPc2pt7CVCUUdQAm7Q9nK-?usp=sharing}.}

\begin{acknowledgments}
  {\bf We would like to thank the referee, Michael~S. Petersen, for their
  thorough report which greatly strengthened the arguments in our work.} We
  would like to thank Greg~L. Bryan, Neal~J. Evans, Drummond~B. Fielding, Keith
  Hawkins, Jason~A.~S. Hunt, Sarah~M.~R. Jeffreson, Kathryn~V. Johnston,
  Peter~M.~W. Kalberla, Jürgen Kerp, Julio~F. Navarro, Dylan Nelson, Suchira
  Sarkar, Joshua~S. Speagle, Martin~D. Weinberg, and Yanfei Zou for helpful
  discussions. A.B. would like to thank Todd Phillips for helpful discussions.
  Some computations in this paper were run on the FASRC Cannon cluster supported
  by the FAS Division of Science Research Computing Group at Harvard University.
  Resources supporting this work were also provided by the NASA High-End
  Computing (HEC) Program through the NASA Advanced Supercomputing (NAS)
  Division at Ames Research Center. This research has made use of the Spanish
  Virtual Observatory (https://svo.cab.inta-csic.es) project funded by
  MCIN/AEI/10.13039/501100011033/ through grant PID2020-112949GB-I00. A.B. was
  supported by the Future Investigators in NASA Earth and Space Science and
  Technology (FINESST) award number 80NSSC20K1536 during the completion of this
  work. E.D. was partially supported by HST grants HST-AR-16363.001 and
  HST-AR-16602.006-A and by NASA Award NASA 80NSSC22K0761. J.Q. acknowledges
  support from NSF grant AST-2008490. L.V.S. is grateful for financial support
  from NASA ATP 80NSSC20K0566, NSF AST 1817233 and NSF CAREER 1945310 grants.
  P.T. acknowledges support from NSF grant AST-1909933, AST-2008490, and NASA
  ATP Grant 80NSSC20K0502. M.V. acknowledges support through NASA ATP
  19-ATP19-0019, 19-ATP19-0020, 19-ATP19-0167, and NSF grants AST-1814053,
  AST-1814259, AST-1909831, AST-2007355 and AST-2107724.
\end{acknowledgments}

@@

\section{Stars Instead of Gas}
In the SMUGGLE model considered in this work, we instantaneously added gas to
the $N$-body system after $1.5\,\textrm{Gyr}$ of evolution. One might wonder if
this sudden change to the potential is responsible for the stable pattern speed
evolution. To test whether this is the case, we added mass to the system in the
same way we did for the SMUGGLE model, but using collisionless particles instead
of gas. The result of this experiment is shown in Fig.~\ref{fig:ps-star}. While
there is an offset compared to the pure $N$-body case, we see that the pattern
speed evolution is broadly consistent with a declining pattern speed. This
indicates that the gas phase is responsible for the stable pattern speed.

\section{Smooth Interstellar Medium}
We performed a simulation of the same disk but with a simpler model of the
interstellar medium \citep{2003MNRAS.339..289S}, closer to standard methods used
in cosmological simulations of galaxy formation and described in more detail in
Section~\ref{sec:methods}. The result of this test is presented in
Fig.~\ref{fig:GFM}. We find that the pattern speed evolution is nearly the same
in this case, and so conclude that our result is not sensitive to the details of
the model for the interstellar medium.

\section{Semi-Analytic Model Parameters}
\label{app:sam}
Our semi-analytic model consisted of a three-component bar-disk-halo system. We
describe here the parameters we chose for these components. The parameters of
the disk and halo were chosen to match closely what we used in our fiducial
simulations. The system can thus be understood as being roughly similar to the
Milky Way, though no careful analysis has been performed to ensure the closest
match possible.
